---
title: "Health Inequalities in England: Deprivation and Life Expectancy Analysis"
author: "Usman Adamu"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE,
                      fig.width = 10,
                      fig.height = 6)
library(fingertipsR)
library(tidyverse)
library(tmap)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
```

## Introduction

Health inequalities remain one of the most persistent challenges facing 
public health in England. Life expectancy, a fundamental measure of 
population health varies substantially across geographic areas and 
socioeconomic groups, reflecting deep-rooted structural inequalities in 
the social determinants of health. The relationship between deprivation 
and life expectancy has been extensively documented, with the Marmot Review 
establishing that life expectancy follows a social gradient: the more 
deprived the area, the shorter the lives of the people who live there 
(Marmot et al., 2010).

This report presents an epidemiological analysis of life expectancy trends 
and health inequalities in England using data from the Office for Health 
Inequalities and Disparities (OHID) Fingertips platform. The analysis 
examines national trends in life expectancy by sex, geographic variation 
across local authorities, the relationship between deprivation and life 
expectancy, and inequality in life expectancy between the most and least 
deprived areas.

### Research Questions

1. How has life expectancy changed in England between 2001 and 2024 
   for males and females?
2. Which local authorities have the highest and lowest life expectancy 
   in England?
3. What is the geographic distribution of life expectancy across 
   English local authorities?
4. Is there a statistically significant relationship between 
   deprivation and life expectancy at local authority level?
5. Does the relationship between deprivation and life expectancy 
   differ by sex?
6. How does inequality in life expectancy between the most and least 
   deprived areas vary over time?
7. How does total life expectancy compare to healthy life expectancy, 
   and how has this gap changed over time?

## Methods

### Study Design

This is an ecological cross-sectional analysis using routinely collected, 
publicly available surveillance data from the OHID Fingertips platform. 
The primary unit of analysis is the local authority (Counties and Unitary 
Authorities from April 2023 (AreaTypeID 502) in England. A longitudinal 
component is incorporated for national trend analyses, using England-level 
aggregated time series data spanning 2001 to 2024. Ecological study designs 
allow population-level patterns and associations to be examined efficiently 
using routine data, though findings cannot be assumed to hold at the 
individual level.^12^

### Data Source

Data were obtained from the OHID Fingertips platform using the 
fingertipsR package in R. Four indicators were extracted at the local 
authority level (Counties and Unitary Authorities from April 2023):

- **Life expectancy at birth** (Indicator 90366)
- **Healthy life expectancy at birth** (Indicator 90362)
- **Deprivation score — IMD 2019** (Indicator 93553)
- **Inequality in life expectancy at birth** (Indicator 92901)

All data are publicly available, aggregate, and anonymised at the area level. 
Data were accessed on 9 April 2026. Life expectancy estimates are presented 
as three-year rolling averages, which reduce year-to-year random variation 
and are the standard formats used by ONS and OHID for local authority 
reporting.^1^

### Statistical Methods

The following analytical approaches were applied sequentially to address 
the seven research questions:

1. **National trend analysis** — Descriptive time series of life expectancy 
   at birth for England by sex from 2001–03 to 2022–24, using three-year 
   rolling average values extracted from the Fingertips platform.
2. **Local authority variation** — Identification of the ten local authorities 
   with the highest and lowest female life expectancy at birth in 2022-24, 
   visualised using ranked bar charts.
3. **Geographic mapping** — Choropleth mapping of female life expectancy 
   across English local authorities using the tmap and rnaturalearth packages, 
   with area boundaries joined to the Fingertips data by local authority name.
4. **Deprivation regression (female)** — Simple linear regression of female 
   life expectancy (outcome) on IMD 2019 deprivation score (exposure) at 
   local authority level, with the coefficient of determination (R²) 
   reported as a measure of explanatory power.
5. **Sex-stratified regression comparison** — Equivalent linear regression 
   models fitted separately for males and females, with regression slopes 
   (β coefficients) compared to quantify differential deprivation effects 
   by sex.
6. **Inequality trend analysis** — Descriptive time series of the 
   life expectancy gap between the most and least deprived areas 
   in England by sex, using the OHID inequality indicator (Indicator 92901).
7. **Healthy life expectancy comparison** — Stacked bar chart comparing 
   years spent in good health versus years spent in poor health by sex 
   across available time periods, derived by subtracting healthy life 
   expectancy from total life expectancy.

All analyses were conducted in R (version 4.3 or later) using the 
tidyverse, fingertipsR, ggplot2, tmap, sf, rnaturalearth, and 
rnaturalearthdata packages. A significance threshold of p < 0.05 
was applied for all statistical tests.

### Ethical Considerations

All data used in this analysis are publicly available, aggregate, and 
fully anonymised at the area level. No individual-level data were accessed 
at any stage of the analysis. In accordance with standard guidance for 
analyses of publicly available, non-identifiable data, no formal ethical 
approval was required for this study.

```{r packages, include=FALSE}
library(fingertipsR)
library(tidyverse)
library(tmap)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
```


```{r load-data, results='hide'}
life_expectancy <- fingertips_data(
  IndicatorID = 90366,
  AreaTypeID = 502)

healthy_life_expectancy <- fingertips_data(
  IndicatorID = 90362,
  AreaTypeID = 502)

deprivation <- fingertips_data(
  IndicatorID = 93553,
  AreaTypeID = 502)

inequality_le <- fingertips_data(
  IndicatorID = 92901,
  AreaTypeID = 502)

# Clean England trend data
le_england <- life_expectancy %>%
  dplyr::filter(AreaName == "England",
                Sex %in% c("Male", "Female"),
                stringr::str_detect(Timeperiod, "-")) %>%
  dplyr::select(Timeperiod, Sex, Value, 
                LowerCI95.0limit, UpperCI95.0limit) %>%
  dplyr::filter(!is.na(Value))

# Clean local authority data - most recent period
le_la <- life_expectancy %>%
  dplyr::filter(AreaType == "Counties & UAs (from Apr 2023)",
                Sex == "Female",
                Timeperiod == "2022 - 24") %>%
  dplyr::select(AreaName, Value, 
                LowerCI95.0limit, UpperCI95.0limit) %>%
  dplyr::filter(!is.na(Value))

# Clean deprivation data
dep_la <- deprivation %>%
  dplyr::filter(AreaType == "Counties & UAs (from Apr 2023)") %>%
  dplyr::select(AreaName, Value) %>%
  dplyr::rename(deprivation_score = Value) %>%
  dplyr::filter(!is.na(deprivation_score))

# Check missing values
cat("Missing values in life expectancy data:", 
    sum(is.na(le_england$Value)), "\n")
cat("Missing values in deprivation data:", 
    sum(is.na(dep_la$deprivation_score)), "\n")
```

## Results

### 1. National Life Expectancy Trend 2001–2024

The chart below shows how life expectancy at birth has changed in 
England over two decades, by sex.

```{r national-trend}
ggplot(le_england, aes(x = Timeperiod, y = Value, 
                        colour = Sex, group = Sex)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_colour_manual(values = c("Male" = "#2196F3",
                                 "Female" = "#FF5722")) +
  labs(title = "Life Expectancy at Birth in England 2001\u20132024",
       subtitle = "By sex, three-year rolling averages",
       x = "Period",
       y = "Life Expectancy (years)",
       colour = "Sex") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 9),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

Life expectancy at birth in England increased steadily in both sexes 
between 2001-03 and 2012-14, consistent with sustained improvements 
in healthcare, medicines, vaccination coverage and socioeconomic 
conditions during this period.^1,2^ However, this improvement stalled 
markedly after 2012-14 and remained largely static until 2018-20,  
a pattern attributed in the literature to the effects of austerity 
policies, stagnating income growth and widening social inequalities 
following the 2008 financial crisis.^3^ A sharp decline in life 
expectancy was observed in both sexes during 2019-21, most likely 
reflecting direct and indirect mortality attributable to the COVID-19 
pandemic, including disruption to routine healthcare services and 
excess deaths among vulnerable populations.^4^ Partial recovery was 
observed from 2021-23 onwards, though life expectancy had not returned 
to the pre-pandemic trajectory by the most recent data point. Throughout 
the study period, female life expectancy consistently exceeded male 
life expectancy by approximately three to four years, consistent with 
established sex differentials in mortality in high-income countries.^5^ 
It should be noted that a proportion of the observed increase in 
recorded life expectancy over time may partly reflect improvements in 
data completeness and reporting quality rather than true biological 
change alone.

### 2. Local Authority Variation in Life Expectancy

The charts below show the local authorities with the highest and 
the lowest female life expectancy at birth in 2022-24.

```{r la-variation}
# Top 10 highest
top_10_le <- le_la %>%
  dplyr::arrange(desc(Value)) %>%
  dplyr::slice_head(n = 10)

# Bottom 10 lowest
bottom_10_le <- le_la %>%
  dplyr::arrange(Value) %>%
  dplyr::slice_head(n = 10)

# Plot top 10
ggplot(top_10_le, aes(x = reorder(AreaName, Value), 
                      y = Value, fill = Value)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = round(Value, 1)),
            hjust = -0.2, size = 4, fontface = "bold") +
  scale_fill_gradient(low = "#B3FFB3", high = "#00AA00") +
  coord_flip() +
  labs(title = "Top 10 Local Authorities: Highest Life Expectancy",
       subtitle = "Female life expectancy at birth, 2022-24",
       x = "",
       y = "Life Expectancy (years)") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11),
    axis.text = element_text(size = 10),
    legend.position = "none"
  )
```

```{r bottom-10}
# Plot bottom 10
ggplot(bottom_10_le, aes(x = reorder(AreaName, -Value),
                          y = Value, fill = Value)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = round(Value, 1)),
            hjust = -0.2, size = 4, fontface = "bold") +
  scale_fill_gradient(low = "#FF0000", high = "#FFB3B3") +
  coord_flip() +
  labs(title = "Top 10 Local Authorities: Lowest Life Expectancy",
       subtitle = "Female life expectancy at birth, 2022-24",
       x = "",
       y = "Life Expectancy (years)") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11),
    axis.text = element_text(size = 10),
    legend.position = "none"
  )
```

Considerable geographic variation in life expectancy exists across 
English local authorities, with a clear North-South gradient apparent 
in the 2022-24 data. Local authorities with the highest life expectancy 
are predominantly concentrated in southern England, while those with 
the lowest are located in northern regions, consistent with the 
established health inequality gradient documented extensively in the 
Marmot Reviews.^3,6^ A female child born in Kensington and Chelsea 
would expect to live approximately eight years longer than her 
counterpart born in Blackpool, a gap that has persisted for decades 
and reflects entrenched socioeconomic inequality rather than individual 
circumstance.^3,6^

This geographic disparity reflects deep-rooted structural differences 
in the social determinants of health, including income, employment, 
housing quality, educational attainment and access to services, 
factors that cluster geographically and compound over time.^3^ It 
should be noted that life expectancy alone provides an incomplete 
picture of population health. Healthy life expectancy, the number 
of years lived in good health, is an equally important measure, as 
it captures the quality as well as the quantity of life. Evidence 
consistently shows that the healthy life expectancy gap between 
deprived and affluent areas is substantially wider than the life 
expectancy gap alone, meaning that residents of the most deprived 
areas not only live shorter lives but also spend a greater proportion of 
those lives in poor health.^7^ This compounds the burden of 
deprivation and underscores the need for upstream policy interventions 
targeting the social determinants of health rather than the downstream 
clinical treatment alone.

### 3. Geographic Distribution of Life Expectancy

The map below shows female life expectancy at birth across English 
local authorities in 2022-24.

```{r geography-map}
# Get England boundaries
england_boundaries <- ne_states(country = "united kingdom",
                                returnclass = "sf") %>%
  dplyr::filter(!region %in% c("Northern Ireland",
                               "West Wales and the Valleys",
                               "East Wales",
                               "Highlands and Islands",
                               "North Eastern",
                               "South Western",
                               "Eastern"))

# Fix name mismatches
le_la_map <- le_la %>%
  dplyr::mutate(AreaName = dplyr::recode(AreaName,
    "Windsor and Maidenhead" = "Royal Borough of Windsor and Maidenhead",
    "County Durham" = "Durham",
    "Bournemouth, Christchurch and Poole" = "Bournemouth",
    "Buckinghamshire UA" = "Buckinghamshire",
    "North Northamptonshire" = "Northamptonshire",
    "West Northamptonshire" = "Northamptonshire",
    "North Yorkshire UA" = "North Yorkshire",
    "Somerset UA" = "Somerset",
    "St. Helens" = "St Helens",
    "Cumberland" = "Cumbria",
    "Westmorland and Furness" = "Cumbria",
    "City of London" = "Greater London",
    "Wirral" = "Merseyside"
  ))

# Join to boundaries
england_map <- england_boundaries %>%
  dplyr::left_join(le_la_map, by = c("name" = "AreaName"))

# Plot map
tm_shape(england_map) +
  tm_polygons(fill = "Value",
              fill.scale = tm_scale_continuous(
                values = "brewer.greens"),
              fill.legend = tm_legend(
                title = "Life Expectancy (years)")) +
  tm_title("Female Life Expectancy at Birth by Local Authority 2022-24") +
  tm_layout(bg.color = "#f5f5f5")
```

The choropleth map confirms the North-South gradient observed in the 
bar charts, with darker green shading, indicating higher life 
expectancy, concentrated predominantly in southern England, 
particularly in London, the South East and the South West. While 
isolated areas of higher life expectancy exist across the country, 
the overall spatial pattern is consistent with evidence identifying 
neighbourhood socioeconomic deprivation as a key driver of geographic 
inequalities in health outcomes in England.^8^ Areas around Liverpool, 
Manchester and the North East show consistently lighter shading, 
reflecting persistently lower life expectancy in these regions. This 
geographic clustering of poor health outcomes in northern areas is 
not coincidental, it reflects decades of structural economic 
disadvantage, industrial decline and underinvestment in social 
infrastructure that has shaped the health of these populations over 
generations.^3,6^

### 4. Deprivation and Life Expectancy — Is There a Relationship?

The scatter plot below examines the relationship between neighbourhood 
deprivation and female life expectancy across English local authorities.

```{r deprivation-join, results='hide'}
le_dep <- le_la %>%
  dplyr::inner_join(dep_la, by = "AreaName")

cat("Number of matched local authorities:", nrow(le_dep), "\n")
```

```{r deprivation-scatter}
ggplot(le_dep, aes(x = deprivation_score, y = Value)) +
  geom_point(colour = "#2196F3", size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", colour = "#FF5722", 
              se = TRUE, linewidth = 1.2) +
  labs(title = "Deprivation Score vs Female Life Expectancy",
       subtitle = "English local authorities, 2022-24",
       x = "Deprivation Score (IMD 2019)",
       y = "Female Life Expectancy at Birth (years)") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )
```

```{r deprivation-regression, results='hide'}
le_model <- lm(Value ~ deprivation_score, data = le_dep)
summary(le_model)
```

Linear regression confirmed a statistically significant negative 
relationship between deprivation score and female life expectancy 
across English local authorities (β = -0.16, p < 0.001, R² = 0.67). 
For every one unit increase in IMD deprivation score, female life 
expectancy decreased by approximately 0.16 years, equivalent to 
nearly two months of life lost per deprivation unit. Deprivation 
score alone explained 66.7% of the variance in female life expectancy 
across local authorities, representing an exceptionally strong 
association for a single predictor in ecological analysis. This 
finding is consistent with the foundational evidence from the Marmot 
Review establishing that life expectancy follows a social gradient 
across England, and with more recent evidence demonstrating that 
socioeconomic deprivation remains the dominant driver of geographic 
health inequalities.^3,6^ The presence of a small number of outlying 
local authorities, those with higher or lower life expectancy than 
predicted by their deprivation score alone, suggests that additional 
factors including local health service quality, population composition, 
migration patterns and environmental conditions may modify the 
relationship between deprivation and mortality in specific localities. 
These findings are consistent with the established evidence that 
place-based deprivation is among the strongest determinants of 
population health in England.^3,8^

### 5. Deprivation and Life Expectancy by Sex

The scatter plots below examine whether the relationship between 
deprivation and life expectancy differ between males and females 
across English local authorities.

```{r le-male, results='hide'}
# Get male life expectancy for latest period
le_la_male <- life_expectancy %>%
  dplyr::filter(AreaType == "Counties & UAs (from Apr 2023)",
                Sex == "Male",
                Timeperiod == "2022 - 24") %>%
  dplyr::select(AreaName, Value) %>%
  dplyr::rename(le_male = Value) %>%
  dplyr::filter(!is.na(le_male))

# Join with deprivation
dep_male <- le_la_male %>%
  dplyr::inner_join(dep_la, by = "AreaName")

# Run regression
male_model <- lm(le_male ~ deprivation_score, data = dep_male)
summary(male_model)
```

```{r combined-scatter}
# Combine male and female data
le_la_female <- le_la %>%
  dplyr::rename(le_value = Value) %>%
  dplyr::mutate(Sex = "Female")

le_la_male_plot <- dep_male %>%
  dplyr::rename(le_value = le_male) %>%
  dplyr::mutate(Sex = "Male")

le_combined <- dplyr::bind_rows(
  le_la_female %>% 
    dplyr::inner_join(dep_la, by = "AreaName") %>%
    dplyr::select(AreaName, le_value, deprivation_score, Sex),
  le_la_male_plot %>%
    dplyr::select(AreaName, le_value, deprivation_score, Sex)
)

ggplot(le_combined, aes(x = deprivation_score, y = le_value,
                         colour = Sex)) +
  geom_point(size = 2.5, alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.2) +
  scale_colour_manual(values = c("Male" = "#2196F3",
                                 "Female" = "#FF5722")) +
  labs(title = "Deprivation vs Life Expectancy by Sex",
       subtitle = "English local authorities, 2022-24",
       x = "Deprivation Score (IMD 2019)",
       y = "Life Expectancy at Birth (years)",
       colour = "Sex") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )
```

Regression modelling confirmed statistically significant negative 
associations between deprivation score and life expectancy in both 
sexes (both p < 0.001). However, the strength and magnitude of this 
relationship differed markedly between males and females. For males, 
deprivation score explained 79.8% of the variance in life expectancy 
(β = -0.191, R² = 0.798), compared to 66.7% for females 
(β = -0.163, R² = 0.667). This indicates that male life expectancy 
is more tightly determined by area-level deprivation than female 
life expectancy, and that males lose approximately 0.03 additional 
years of life per deprivation unit compared to females.^3,6^

The combined scatter plot illustrates these differential gradients 
clearly, while females maintain higher life expectancy than males 
at every level of deprivation, the male regression line is visibly 
steeper, indicating that the penalty of living in a deprived area 
is greater for men than for women. This pattern is consistent with 
evidence that men in deprived areas face disproportionately higher 
mortality from cardiovascular disease, suicide, alcohol-related 
conditions and occupational hazards, causes of death that are 
strongly socially patterned and more prevalent among men.^9^ These 
findings reinforce the case for gender-sensitive public health 
strategies in deprived areas, targeting the specific causes of 
male excess mortality alongside broader deprivation reduction 
interventions.^3,9^

### 6. Inequality in Life Expectancy Across England

```{r inequality}
inequality_clean <- inequality_le %>%
  dplyr::filter(AreaName == "England",
                !is.na(Value)) %>%
  dplyr::select(Timeperiod, Sex, Value) %>%
  dplyr::filter(Sex %in% c("Male", "Female"),
                stringr::str_detect(Timeperiod, "-"))

ggplot(inequality_clean, aes(x = Timeperiod, y = Value,
                              colour = Sex, group = Sex)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_colour_manual(values = c("Male" = "#2196F3",
                                 "Female" = "#FF5722")) +
  labs(title = "Inequality in Life Expectancy at Birth in England",
       subtitle = "Gap between most and least deprived areas, by sex",
       x = "Period",
       y = "Gap in Life Expectancy (years)",
       colour = "Sex") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 9),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

The inequality in life expectancy between the most and least deprived 
areas of England have remained persistently high throughout the study 
period, showing little meaningful improvement over two decades despite 
national policy commitments to reducing health inequalities.^3,6^ 
Inequality was consistently greater among males than females, 
reflecting the combined effects of higher male mortality rates in 
deprived areas and greater male susceptibility to deprivation-related 
causes of death, including cardiovascular disease, substance misuse 
and occupational hazards.^9^ A notable widening of the inequality gap 
was observed following 2020, consistent with evidence that the 
COVID-19 pandemic disproportionately affected mortality in deprived 
areas, with death rates in the most deprived quintile of local 
authorities exceeding those in the least deprived by over 50% during 
the first wave.^10^ This post-pandemic widening suggests that existing 
health inequalities were not merely preserved but actively exacerbated 
by the pandemic and its associated disruption to healthcare services, 
employment and social support, a finding with significant implications 
for health inequality policy in the recovery period. The persistence 
of this gap over two decades underscores the inadequacy of the downstream 
health interventions alone and points to the need for sustained 
upstream action on the structural determinants of health, including 
income, housing, employment and education.^3,11^

### 7. Life Expectancy vs Healthy Life Expectancy

The chart below compares total life expectancy with healthy life 
expectancy at birth in England over time, highlighting the years 
spent in poor health.

```{r healthy-le, results='hide'}
# Get healthy life expectancy England trend
hle_england <- healthy_life_expectancy %>%
  dplyr::filter(AreaName == "England",
                Sex %in% c("Male", "Female"),
                stringr::str_detect(Timeperiod, "-"),
                !is.na(Value)) %>%
  dplyr::select(Timeperiod, Sex, Value) %>%
  dplyr::rename(healthy_le = Value)

# Get total life expectancy
le_england_plot <- le_england %>%
  dplyr::rename(total_le = Value) %>%
  dplyr::select(Timeperiod, Sex, total_le)

# Join both
le_comparison <- le_england_plot %>%
  dplyr::inner_join(hle_england, by = c("Timeperiod", "Sex")) %>%
  dplyr::mutate(years_poor_health = total_le - healthy_le)

# Diagnostic checks — suppressed from output
head(le_comparison)
nrow(le_comparison)
```

```{r healthy-le-chart}
# Reshape for plotting
le_long <- le_comparison %>%
  dplyr::select(Timeperiod, Sex, healthy_le, years_poor_health) %>%
  tidyr::pivot_longer(cols = c(healthy_le, years_poor_health),
                      names_to = "measure",
                      values_to = "years") %>%
  dplyr::mutate(measure = dplyr::recode(measure,
    "healthy_le" = "Years in Good Health",
    "years_poor_health" = "Years in Poor Health"
  ))

ggplot(le_long, aes(x = Timeperiod, y = years,
                    fill = measure, group = interaction(Sex, measure))) +
  geom_bar(stat = "identity") +
  facet_wrap(~Sex) +
  scale_fill_manual(values = c("Years in Good Health" = "#4CAF50",
                               "Years in Poor Health" = "#FF5722")) +
  labs(title = "Life Expectancy vs Healthy Life Expectancy in England",
       subtitle = "Years in good and poor health by sex, 2011-2024",
       x = "Period",
       y = "Years",
       fill = "") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 8),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom",
    strip.text = element_text(size = 12, face = "bold")
  )
```

The stacked bar chart reveals a paradox at the heart of female 
longevity in England. Despite living approximately four years longer 
than males, females spend a greater absolute number of years in poor 
health — approximately 19 to 21 years compared to 16 to 17 years 
for males throughout the study period. This means that much of the 
female longevity advantage is spent in a state of ill health, 
disability or reduced functioning rather than in full health.^7^

The years spent in poor health remained remarkably stable across 
both sexes throughout the study period, the orange band shows 
virtually no change from 2011-13 to 2021-23, indicating that 
while total life expectancy has fluctuated, the proportion of life 
spent in poor health has not improved meaningfully over more than 
a decade. A marginal increase in female years in poor health was 
observed in the most recent periods from 2020-22 onwards, 
consistent with evidence that the COVID-19 pandemic increased the 
burden of long-term conditions and disability, particularly among 
older women.^4^

These findings have significant implications for NHS planning and 
commissioning. The persistent gap between total and healthy life 
expectancy, representing approximately one fifth of all years 
lived in England, represents a substantial and largely 
unaddressed burden of chronic disease, disability and social care 
need. Policies focused solely on extending life expectancy without 
addressing the quality of those additional years risk increasing 
the demand for health and social care services without improving 
population wellbeing.^3^ Healthy life expectancy should be adopted 
as a primary outcome measure in public health planning alongside 
total life expectancy.^6^

## Limitations

Several limitations should be considered when interpreting the findings 
of this analysis.

**Ecological fallacy.** All analyses were conducted at the local authority 
level. The associations observed between area-level deprivation and life 
expectancy cannot be assumed to hold at the individual level, a 
phenomenon known as the ecological fallacy.^12^ Individual-level data 
would be required to confirm whether the associations identified here 
operate through the same mechanisms at the person level.

**Deprivation data currency.** The deprivation measure used (IMD 2019) 
predates the COVID-19 pandemic and the subsequent cost-of-living crisis, 
both of which are likely to have materially altered the socioeconomic 
profile of many local authorities. The true contemporary relationship 
between deprivation and life expectancy may therefore differ from that 
estimated here, and the IMD 2019 index may underestimate current 
deprivation in areas most severely affected by the post-pandemic economic 
deterioration.

**Univariable regression.** The regression models in Sections 4 and 5 
include only the deprivation score as a predictor. Life expectancy is 
determined by multiple interacting factors, including ethnicity, rurality, 
access to healthcare, environmental quality and population age structure. 
The high R² values observed may partly reflect the correlation between 
deprivation and these unmeasured variables rather than a direct 
deprivation-only effect. Multivariable modelling would be required to 
isolate the independent contribution of deprivation.

**Geographic boundary mismatches.** Joining Fingertips local authority 
data to rnaturalearth boundaries required manual name recoding due to 
inconsistencies in naming conventions. Some local authority boundaries 
used by OHID do not have exact equivalents in the rnaturalearth dataset, 
necessitating approximations (for example, mapping Cumberland and 
Westmorland and Furness to the legacy Cumbria boundary). This may have 
introduced minor inaccuracies in the choropleth map.

**Sex-limited geographic analysis.** The geographic bar charts and 
choropleth map presents female life expectancy only. Male geographic 
variation across local authorities was not visualised, which means 
spatial patterns of male life expectancy inequality cannot be directly 
assessed from this analysis.

**Causality.** This is a cross-sectional ecological analysis. No causal 
inferences can be drawn from the associations identified. Observed 
relationships between deprivation and life expectancy are consistent 
with a causal pathway, but cannot confirm one.

**Data updates.** Fingertips data are periodically revised by OHID. The 
specific values reported here reflect the data available at the time of 
access (9 April 2026) and may differ from values available at other times 
points.

## Conclusion

This report examined health inequalities in England through the lens 
of life expectancy and deprivation, using real data from the OHID 
Fingertips platform covering local authorities across England. The 
analysis extended beyond a simple description of trends to formally 
quantify the relationship between deprivation and life expectancy 
by sex, compare total and healthy life expectancy, and examine the 
persistence of inequality over two decades.

### Key Findings

**1. Stalling and reversal of life expectancy gains**

Life expectancy in England increased steadily from 2001 to 2012-14 
before stalling markedly, a pattern attributed to austerity 
policies and widening socioeconomic inequalities following the 2008 
financial crisis.^3^ The COVID-19 pandemic caused a sharp decline 
in life expectancy in both sexes from 2019-21, with only partial 
recovery observed by 2022-24. Female life expectancy consistently 
exceeded male life expectancy throughout the study period by 
approximately three to four years.^5^

**2. Stark geographic inequality**

A clear North-South gradient exists in life expectancy across 
English local authorities, with local authorities in southern 
England, particularly London and the South East, consistently 
showing higher life expectancy than those in northern regions. 
A female child born in the local authority with the highest life 
expectancy would expect to live approximately eight years longer 
than her counterpart born in the local authority with the lowest,
a gap reflecting entrenched structural inequalities in the 
social determinants of health.^3,6^

**3. Deprivation as the dominant predictor — with greater impact on males**

Linear regression confirmed that deprivation score explained 
66.7% of the variance in female life expectancy (β = -0.163, 
p < 0.001) and 79.8% of the variance in male life expectancy 
(β = -0.191, p < 0.001) across local authorities. Males lose 
approximately 0.03 additional years of life per deprivation unit 
compared to females, confirming that the penalty of deprivation 
is greater for men than for women. For every one unit increase 
in IMD deprivation score, female life expectancy decreases by 
0.16 years and male life expectancy by 0.19 years.^3,9^

**4. The healthy life expectancy paradox**

Despite living longer than males, females spend approximately 
19 to 21 years in poor health, compared to 16 to 17 years for 
males. The years spent in poor health remained stubbornly stable 
across both sexes from 2011 to 2023, with no meaningful 
improvement over more than a decade. A marginal increase in 
female years in poor health were observed post-2020, consistent 
with the long-term health impact of the COVID-19 pandemic.^4,7^ 
These findings highlight that extending life expectancy without 
improving healthy life expectancy risks increasing the burden 
of chronic disease and social care needs without improving 
population wellbeing.

**5. Persistent and widening inequality**

Inequality in life expectancy between the most and least deprived 
areas have remained persistently high over two decades, with little 
meaningful improvement despite national policy commitments to 
reducing health inequalities.^3,6^ A notable widening of the gap 
was observed following 2020, consistent with evidence that the 
COVID-19 pandemic disproportionately increased mortality in the 
most deprived areas, actively exacerbating pre-existing inequalities 
rather than merely preserving them.^10,11^

### Recommendations

The findings of this analysis support the following evidence-based 
recommendations for public health policy and practice.

**1. Prioritise upstream action on the social determinants of health.**
Given that deprivation alone explains up to 79.8% of the variance in 
local authority life expectancy, clinical interventions alone are 
insufficient to close the health gap. National and local government 
should prioritise sustained investment in income support, affordable 
housing, employment, and educational attainment, particularly in 
persistently deprived northern local authorities.^3,6^

**2. Adopt healthy life expectancy as a co-primary outcome in public 
health planning.**
Total life expectancy masks the substantial burden of years lived in 
poor health. NHS England, integrated care boards and local authorities 
should formally adopt healthy life expectancy as a co-primary planning 
outcome alongside total life expectancy, ensuring that commissioning 
decisions are evaluated against both quantity and quality of life.^6^

**3. Implement targeted post-pandemic recovery strategies in the most 
deprived areas.**
The widening of the life expectancy inequality gap post-2020 signals 
that the COVID-19 pandemic disproportionately set back health outcomes 
in already disadvantaged communities. Dedicated, place-based recovery 
programmes, modelled on Marmot Place initiatives, should be 
prioritised in those local authorities where the gap widened most 
sharply.^11^

**4. Develop gender-sensitive strategies targeting male excess mortality 
in deprived areas.**
The steeper deprivation gradient observed in males, and the 
disproportionate contribution of cardiovascular disease, suicide, 
alcohol-related conditions and occupational hazards to male excess 
mortality, requires sex-disaggregated public health strategies. 
Programmes targeting male health-seeking behaviour, mental health, 
and substance misuse in deprived areas should be developed alongside 
broader structural interventions.^9^

**5. Strengthen surveillance and monitoring of healthy life expectancy 
inequality.**
Current national monitoring focuses predominantly on total life 
expectancy inequality. OHID and NHS England should expand routine 
reporting to include the healthy life expectancy gap between the most 
and least deprived areas, enabling a more complete picture of health 
inequality trends to inform policy.^13^

**6. Commission further analytical work using multivariable and 
individual-level data.**
The ecological, univariable approach used in this analysis, while 
suitable for identifying population-level patterns, cannot establish 
causality or isolate the independent effect of deprivation from 
confounders. Multivariable analyses using linked individual-level 
datasets, such as the Clinical Practice Research Datalink or the 
Health Survey for England would strengthen the evidence base for 
targeted interventions.

### Data Reproducibility

This analysis was conducted entirely in R using the fingertipsR package 
to access OHID Fingertips data programmatically, ensuring that the data 
extraction process is transparent, documented, and reproducible. All 
indicator IDs, area type codes, and data processing steps are specified 
in the code. The complete analysis code is available at 
https://usman-adamu.github.io/health-inequalities-england/ for independent
verification and replication. Readers should note that OHID periodically
updates Fingertips data; results may therefore differ marginally if the analysis 
is reproduced at a later date using the same indicator IDs.

## References

1. Office for National Statistics. National life tables — England 
   [Internet]. Newport: ONS; 2024 [cited 2026 Apr 9]. Available 
   from: https://www.ons.gov.uk

2. Public Health England. Health profile for England [Internet]. 
   London: PHE; 2019 [cited 2026 Apr 9]. Available from: 
   https://fingertips.phe.org.uk

3. Marmot M, Allen J, Boyce T, Goldblatt P, Morrison J. Health 
   equity in England: the Marmot Review 10 years on. London: 
   Institute of Health Equity; 2020.

4. Office for National Statistics. Deaths registered weekly in 
   England and Wales, provisional: 2020 [Internet]. Newport: 
   ONS; 2021 [cited 2026 Apr 9]. Available from: 
   https://www.ons.gov.uk

5. Austad SN, Bartke A. Sex differences in longevity and in 
   responses to anti-aging interventions: a mini-review. 
   Gerontology. 2016;62(1):40–6.

6. Marmot M, Allen J, Goldblatt P, Boyce T, McNeish D, 
   Grady M. Fair society, healthy lives: the Marmot Review. 
   London: Institute of Health Equity; 2010.

7. Zaninotto P, Falaschetti E, Sacker A. Age trajectories of 
   quality of life among older adults: results from the English 
   Longitudinal Study of Ageing. Qual Life Res. 
   2009;18(10):1301–9.

8. Daras K, Tsakiridi A, Corcoran R, Gabbay M. Unravelling the 
   dynamics of mental health inequalities in England: a 12-year 
   nationwide longitudinal spatial analysis of recorded 
   depression prevalence. SSM Popul Health. 2024;26:101669.

9. Baker P, Dworkin SL, Teng S, Banks I, Shand T, Yamey G. 
   The men's health gap: men must be included in the global 
   health equity agenda. Bull World Health Organ. 
   2014;92(8):618–20.

10. Bambra C, Riordan R, Ford J, Matthews F. The COVID-19 
    pandemic and health inequalities. J Epidemiol Community 
    Health. 2020;74(11):964–8.

11. Institute of Health Equity. Build back fairer: the 
    COVID-19 Marmot Review. London: IHE; 2020.

12. House of Commons Library. Health inequalities: income 
    deprivation and north/south divides [Internet]. London: 
    UK Parliament; 2019 [cited 2026 Apr 9]. Available from: 
    https://commonslibrary.parliament.uk

13. Health Foundation. Quantifying health inequalities in 
    England [Internet]. London: The Health Foundation; 2022 
    [cited 2026 Apr 9]. Available from: 
    https://www.health.org.uk

14. Wickham H, Averick M, Bryan J, Chang W, McGowan LD, 
    François R, et al. Welcome to the tidyverse. J Open Source 
    Softw. 2019;4(43):1686.
